---
title: "Forest Ownership Dynamics"
subtitle: "Estimated Changeds in Land Area by Condition Status and Forest Ownership"
author:
  - Brett J. Butler^[USDA Forest Service, brett.butler2@usda.gov]
output: html_document
---

This report documents the methods used to retrieve, summarize, and analyze data for the "Forest Ownership Dynamics" article (*DOI to be added*).

# General set up
```{r General Set up, warning = F, message = F}
library(tidyverse)
library(RODBC)
library(usmap)
source("../R/get_evalgrps.R")
source("../R/get_counties.R")
source("../R/get_estimates.R")
source("../R/get_base_estimates.R")
source("../R/get_plot.R")
ref.cnty <- read_csv("../DATA/_REF_COUNTY.csv", col_type = "fffcff")
ref.state <- read_csv("../DATA/_REF_STATE.csv", col_type = "fffcff")
ref.foresttype <- read_csv("../DATA/_REF_FOREST_TYPE.csv", col_type = "fcfc")
# cbp <- c("#009E73", "#F0E442", "#56B4E9", "#E69F00", "#0072B2", "#D55E00", "#CC79A7", "#999999", "#000000") 
```

## Get EVAL_GRPs
An Evaluation Group (EVAL_GRP) is a unique value in the FIA database that identifies the FIA NFI sample used to generate estimates (Burrill et al. 2024). The specific EVAL_GRPs used for the forest area estimates generated from this script are listed in the Appendix of this report. 

```{r EVALIDs}
evalgrps <- list(evalgrps.2023 = read_csv("../DATA/EVALGRPS_2023.csv", col_types = "cfcii"),
                 evalgrps.2018 = read_csv("../DATA/EVALGRPS_2018.csv", col_types = "cfcii"),
                 evalgrps.2013 = read_csv("../DATA/EVALGRPS_2013.csv", col_types = "cfcii"))
```

Alaska, Oklahoma, and Texas are divided by FIA NFI into sub-state regions, defined by groups of counties, for sampling and estimation. In addition, an issue with the 2005 sample in interior Alaska where a large acreage of wildernss area was not inventoried, had to be addressed (i.e., excluding an estimaion unit set up to explicitly deal with this issue).

```{r EVAL_GRP Add Counties for AK OK TX}
evalgrps <- lapply(evalgrps, function(x) { x %>%
    mutate(COUNTIES = case_when(STATECD_NWOS == "2.1" ~ get_counties("2.1"),
                                STATECD_NWOS == "40.1" ~ get_counties("40.1"),
                                STATECD_NWOS == "40.2" ~ get_counties("40.2"),
                                STATECD_NWOS == "48.1" ~ get_counties("48.1"),
                                STATECD_NWOS == "48.2" ~ get_counties("48.2"),
                                .default = "IS NOT NULL"))})
evalgrps <- lapply(evalgrps, function(x) { x %>%
    mutate(ESTN_UNIT = case_when(STATECD_NWOS == "2.1" ~ "<> 4",
                                 .default = "IS NOT NULL"))})
```

## Summaries of EVAL_GRPs
```{r EVAL_GRP Summaries}
evalgrp.summary <- bind_rows(evalgrps$evalgrps.2023 %>% mutate(YEAR = 2023),
                             evalgrps$evalgrps.2018 %>% mutate(YEAR = 2018),
                             evalgrps$evalgrps.2013 %>% mutate(YEAR = 2013)) %>%
  select(STATECD = STATECD_NWOS, YEAR, START_INVYR, END_INVYR) %>%
  pivot_wider(names_from = YEAR, values_from = c(START_INVYR, END_INVYR))
evalgrp.summary
summary(evalgrp.summary)
```

# Forest Trends

## Get Base Estimates
```{r Get Base Estimates}
# base.estimates.2023 <- bind_rows(mapply(get_base_estimates,
#                          evalgrp = (evalgrps$evalgrps.2023 %>% pull(EVAL_GRP)),
#                          statecd.nwos = (evalgrps$evalgrps.2023 %>% pull(STATECD_NWOS)),
#                          cnty = (evalgrps$evalgrps.2023 %>% pull(COUNTIES)),
#                          SIMPLIFY = F)) %>% mutate(YEAR = 2023)
# base.estimates.2013 <- bind_rows(mapply(get_base_estimates,
#                          evalgrp = (evalgrps$evalgrps.2013 %>% pull(EVAL_GRP)),
#                          statecd.nwos = (evalgrps$evalgrps.2013 %>% pull(STATECD_NWOS)),
#                          cnty = (evalgrps$evalgrps.2013 %>% pull(COUNTIES)),
#                          eu = (evalgrps$evalgrps.2013 %>% pull(ESTN_UNIT)),
#                          SIMPLIFY = F)) %>% mutate(YEAR = 2013)
# base.estimates <- bind_rows(base.estimates.2023,
#                             base.estimates.2013)
# write_csv(base.estimates, "../DATA/BASE_ESTIMATES.csv")
base.estimates <- read_csv("../DATA/BASE_ESTIMATES.csv", col_types = "_fd__d___fi") %>%
  separate_wider_delim(STATECD_NWOS, ".", too_many = "drop", names = "STATECD", cols_remove = F) %>%
  mutate(STATECD = as.factor(STATECD)) %>%
  rename(ACRES = ESTIMATE, ACRES_VARIANCE = VAR_OF_ESTIMATE) %>%
  mutate(COND_STATUS_CD = case_match(COND_STATUS_CD,
                                     "1" ~ "FOREST",
                                     "2" ~ "NONFOREST",
                                     c("3", "4") ~ "WATER",
                                     .default = COND_STATUS_CD)) %>%
  group_by(STATECD, STATECD_NWOS, YEAR, COND_STATUS_CD) %>%
  summarize(across(c(ACRES, ACRES_VARIANCE), sum), .groups = "drop") %>%
  select(STATECD, STATECD_NWOS, YEAR, COND_STATUS_CD, ACRES, ACRES_VARIANCE) %>%
  arrange(STATECD_NWOS, COND_STATUS_CD, YEAR)
```

```{r Forest Trends}
base.estimates %>%
  filter(COND_STATUS_CD == "FOREST") %>%
  group_by(YEAR) %>%
  summarize(across(c(ACRES, ACRES_VARIANCE), sum), .groups = "drop") %>%
  mutate(AC = ACRES / 1e6, 
         AC_SE = sqrt(ACRES_VARIANCE) / 1e6) %>%
  select(YEAR, AC, AC_SE) %>%
  pivot_wider(names_from = YEAR, values_from = c(AC, AC_SE)) %>%
  mutate(DIFF = AC_2023 - AC_2013,
         DIFF_PERC = (DIFF / AC_2013) * 100,
         DIFF_ANNUAL = DIFF / 10,
         DIFF_ANNUAL_PERC = (DIFF_ANNUAL / AC_2013) * 100) %>%
  mutate(across(AC_2013:AC_SE_2023, ~ round(.x, 1)),
         across(DIFF:DIFF_ANNUAL_PERC, ~ round(.x, 2)))
```

```{r State Totals}
forestarea.state.total <- base.estimates %>%
  filter(COND_STATUS_CD == "FOREST") %>%
  group_by(YEAR, STATECD) %>%
  summarize(across(c(ACRES, ACRES_VARIANCE), sum), .groups = "drop") %>%
  mutate(AC = ACRES / 1e6, AC_SE = sqrt(ACRES_VARIANCE) / 1e6) %>%
  select(YEAR, STATECD, AC, AC_SE) %>%
  pivot_wider(names_from = YEAR, values_from = c(AC, AC_SE)) %>%
  mutate(DIFF = AC_2023 - AC_2013,
         DIFF_PERC = (DIFF / AC_2013) * 100,
         DIFF_ANN = DIFF / 10,
         DIFF_ANN_PERC = (DIFF_ANN / AC_2013) * 100,
         DIFF_ANN_PERC_CAT = cut(DIFF_ANN_PERC, 
                                 breaks = c(-Inf, -0.25, -0.05, 0.05, 0.25, Inf),
                                 labels = c("<-0.25", "-0.25 to -0.06", "0.05 to -0.05",
                                            "0.06 to 0.25", ">0.25")),
         DIFF_ANN_PERC_CAT = factor(DIFF_ANN_PERC_CAT, 
                                    levels = c(">0.25", "0.06 to 0.25", "0.05 to -0.05",
                                               "-0.25 to -0.06","<-0.25")),
         fips = STATECD) 
forestarea.state.total %>% count(DIFF_ANN_PERC_CAT)
plot_usmap("state", include = ref.state %>% distinct(STATE_FIA_ABB) %>% pull(),
           data = forestarea.state.total, values = "DIFF_ANN_PERC_CAT") +
  scale_fill_manual(values = rev(c("#a6611a", "#dfc27d", "#f5f5f5", "#80cdc1", "#018571")), 
                    name = "Percentage\nAnnual\nChange") +
  theme(legend.position = "right")
# ggsave("../FIGURES/ANNUAL_CHANGE_MAP.pdf", width = 6, height = 3)
# ggsave("../FIGURES/ANNUAL_CHANGE_MAP.png", width = 6, height = 3)
```

# Forest Transitions

```{r get_plot}
# plots <- bind_rows(mapply(get_plot,
#                           evalgrp = (evalgrps$evalgrps.2023 %>% pull(EVAL_GRP)),
#                           statecd.nwos = (evalgrps$evalgrps.2023 %>% pull(STATECD_NWOS)),
#                           cnty = (evalgrps$evalgrps.2023 %>% pull(COUNTIES)),
#                           SIMPLIFY = F))
# write_csv(plots, "../DATA/PLOTS.csv")
plots <- read_csv("../DATA/PLOTS.csv", col_types = "ciiiidffffddf") %>%
  separate_wider_delim(STATECD_NWOS, ".", too_many = "drop", names = "STATECD", cols_remove = F) %>%
  mutate(STATECD = as.factor(STATECD),
         COND_STATUS_CD = case_match(COND_STATUS_CD,
                                     "1" ~ "FOREST",
                                     "2" ~ "NONFOREST",
                                     c("3", "4") ~ "WATER",
                                     .default = COND_STATUS_CD),
         PREV_COND_STATUS_CD = case_match(PREV_COND_STATUS_CD,
                                          "1" ~ "FOREST",
                                          "2" ~ "NONFOREST",
                                          c("3", "4") ~ "WATER",
                                          "5" ~ "NONSAMPLED",
                                          .default = PREV_COND_STATUS_CD),
         across(c(PRESNFCD, PREV_PRESNFCD),
                ~ case_match(.x,
                             c("10", "11", "12", "13") ~ "AGRICULTURE",
                             c("20") ~ "RANGELAND",
                             c("30", "31", "32") ~ "DEVELOPED",
                             c("16", "17", "40", "41", "42", "43", "45") ~ "OTHER",
                             .default = .x)),
         LU = if_else(COND_STATUS_CD == "NONFOREST", PRESNFCD, COND_STATUS_CD),
         PREV_LU = if_else(PREV_COND_STATUS_CD == "NONFOREST", PREV_PRESNFCD, PREV_COND_STATUS_CD))
```
## Average Remeasurement by EVAL_GRP

```{r Remeasurement Periods}
remper <- plots %>%
  group_by(STATECD_NWOS) %>%
  summarize(REMPER = mean(REMPER, na.rm = T))
```

```{r get_estimates}
# estimates <- bind_rows(mapply(get_estimates,
#                               evalgrp = (evalgrps$evalgrps.2023 %>% pull(EVAL_GRP)),
#                               statecd.nwos = (evalgrps$evalgrps.2023 %>% pull(STATECD_NWOS)),
#                               cnty = (evalgrps$evalgrps.2023 %>% pull(COUNTIES)),
#                               SIMPLIFY = F))
# write_csv(estimates, "../DATA/ESTIMATES.csv")
estimates <- read_csv("../DATA/ESTIMATES.csv", col_types = "_ffffffffffffffd__d___f") %>%
  separate_wider_delim(STATECD_NWOS, ".", too_many = "drop", names = "STATECD", cols_remove = F) %>%
  rename(ACRES = ESTIMATE, ACRES_VARIANCE = VAR_OF_ESTIMATE) %>%
  mutate(COND_STATUS_CD = case_match(COND_STATUS_CD,
                                     "1" ~ "FOREST",
                                     "2" ~ "NONFOREST",
                                     c("3", "4") ~ "WATER",
                                     .default = COND_STATUS_CD),
         PREV_COND_STATUS_CD = case_match(PREV_COND_STATUS_CD,
                                          "1" ~ "FOREST",
                                          "2" ~ "NONFOREST",
                                          c("3", "4") ~ "WATER",
                                          "5" ~ "NONSAMPLED",
                                          .default = PREV_COND_STATUS_CD),
         across(c(PRESNFCD, PREV_PRESNFCD),
                ~ case_match(.x,
                             c("10", "11", "12", "13") ~ "AGRICULTURE",
                             c("20") ~ "RANGELAND",
                             c("30", "31", "32") ~ "DEVELOPED",
                             c("16", "17", "40", "41", "42", "43", "45") ~ "OTHER",
                             .default = .x)),
         across(c(OWNCD, PREV_OWNCD),
                ~ case_match(.x,
                             c("11", "12", "13", "21",
                               "22", "23", "24", "25") ~ "FEDERAL",
                             c("31") ~ "STATE",
                             c("32", "33") ~ "LOCAL" ,
                             c("41", "42", "43") ~ "CORPORATE",
                             c("44") ~ "TRIBAL",
                             c("45") ~ "FAMILY"))) %>%
  mutate(LU = if_else(COND_STATUS_CD == "FOREST", "FOREST", PRESNFCD),
         PREV_LU = if_else(PREV_COND_STATUS_CD == "FOREST", "FOREST", PREV_PRESNFCD)) %>%
  group_by(STATECD, STATECD_NWOS, COND_STATUS_CD, PREV_COND_STATUS_CD, 
           PRESNFCD, PREV_PRESNFCD, LU, PREV_LU,
           OWNCD, PREV_OWNCD, 
           RESERVCD, PREV_RESERVCD, FORTYPCD, PREV_FORTYPCD,
           STDORGCD, PREV_STDORGCD, SITECLCD, PREV_SITECLCD) %>%
  summarize(across(c(ACRES, ACRES_VARIANCE), sum), .groups = "drop") %>%
  left_join(remper %>%
              group_by(STATECD_NWOS) %>%
              summarize(REMPER = mean(REMPER, na.rm = T)), 
            by = join_by(STATECD_NWOS)) %>%
  mutate(STATECD_NWOS = factor(STATECD_NWOS))
```

## Check Total Forest Acreage 
```{r Forest Acreage Check}
left_join(estimates %>%
            filter(COND_STATUS_CD == "FOREST") %>%
            group_by(STATECD_NWOS) %>%
            summarize(ACRES = sum(ACRES)),
          base.estimates %>%
            filter(YEAR == 2023 & COND_STATUS_CD == "FOREST") %>%
            group_by(STATECD_NWOS) %>%
            summarize(ACRES_BASE = sum(ACRES))) %>%
  mutate(DIFF = (ACRES - ACRES_BASE) / 1e6)
```


## Adjust for Missingness
For the land use transition matrix, missing data related to plots that were not remeasured, not sampled at $t_1$, and missing nonforest type needed to be adjusted for. The remeaurement issue, and related missing nonforest type issue, are primarily an issue in the Intermountain West states and West Texas.

```{r Assess Missingness}
estimates %>% count(COND_STATUS_CD) 
estimates %>% count(PREV_COND_STATUS_CD)

estimates %>% filter(COND_STATUS_CD == "NONFOREST") %>% count(PRESNFCD)
estimates %>% filter(PREV_COND_STATUS_CD == "NONFOREST") %>% count(PREV_PRESNFCD)

estimates %>% filter(COND_STATUS_CD == "NONFOREST") %>% count(LU)
estimates %>% filter(PREV_COND_STATUS_CD == "NONFOREST") %>% count(PREV_LU)
```

### Adjust for Not Remeasured
```{r T_1 NONSAMPLED ADJUST}
nr.adj <- left_join(estimates %>%
                      group_by(STATECD_NWOS) %>%
                      summarize(ACRES_TOTAL = sum(ACRES), .groups = "drop"),
                    estimates %>%
                      filter(!is.na(PREV_COND_STATUS_CD)) %>%
                      group_by(STATECD_NWOS) %>%
                      summarize(ACRES_REMEASURED = sum(ACRES), .groups = "drop"),
                    by = join_by(STATECD_NWOS)) %>%
  mutate(NR_ADJ = ACRES_TOTAL / ACRES_REMEASURED) %>%
  select(STATECD_NWOS, NR_ADJ)
estimates <- estimates %>%
  filter(!is.na(PREV_COND_STATUS_CD)) %>%
  left_join(nr.adj, 
            by = join_by(STATECD_NWOS)) %>%
  mutate(ACRES = if_else(is.na(NR_ADJ), ACRES, ACRES * NR_ADJ)) %>%
  select(-NR_ADJ)
```

### Adjust for T_1 NONSAMPLED
```{r T_1 NONSAMPLED ADJUST}
ns.adj <- left_join(estimates %>%
                      group_by(STATECD_NWOS) %>%
                      summarize(ACRES_TOTAL = sum(ACRES), .groups = "drop"),
                    estimates %>%
                      filter(PREV_COND_STATUS_CD != "NONSAMPLED") %>%
                      group_by(STATECD_NWOS) %>%
                      summarize(ACRES_SAMPLED = sum(ACRES), .groups = "drop"),
                    by = join_by(STATECD_NWOS)) %>%
  mutate(NS_ADJ = ACRES_TOTAL / ACRES_SAMPLED) %>%
  select(STATECD_NWOS, NS_ADJ)
estimates <- estimates %>%
  filter(!(PREV_COND_STATUS_CD == "NONSAMPLED")) %>%
  left_join(ns.adj, 
            by = join_by(STATECD_NWOS)) %>%
  mutate(ACRES = if_else(is.na(NS_ADJ), ACRES, ACRES * NS_ADJ)) %>%
  select(-NS_ADJ)
```

### Adjust for PRESNFCD and PREV_PRESNFCD Missingness
```{r}
nf.adj <- left_join(estimates %>%
                      filter(COND_STATUS_CD == "NONFOREST") %>%
                      group_by(STATECD_NWOS) %>%
                      summarize(ACRES_NONFOREST = sum(ACRES), .groups = "drop"),
                    estimates %>%
                      filter(COND_STATUS_CD == "NONFOREST" & !is.na(PRESNFCD)) %>%
                      group_by(STATECD_NWOS) %>%
                      summarize(ACRES_NONFOREST_NOT_MISSING = sum(ACRES), .groups = "drop"),
                    by = join_by(STATECD_NWOS)) %>%
  mutate(COND_STATUS_CD = "NONFOREST", 
         NF_ADJ = ACRES_NONFOREST / ACRES_NONFOREST_NOT_MISSING) %>%
  select(STATECD_NWOS, COND_STATUS_CD, NF_ADJ)
estimates <- estimates %>%
  filter(!(COND_STATUS_CD == "NONFOREST" & is.na(PRESNFCD))) %>%
  left_join(nf.adj, 
            by = join_by(STATECD_NWOS, COND_STATUS_CD)) %>%
  mutate(ACRES = if_else(is.na(NF_ADJ), ACRES, ACRES * NF_ADJ)) %>%
  select(-NF_ADJ)

prev.nf.adj <- left_join(estimates %>%
                           filter(PREV_COND_STATUS_CD == "NONFOREST") %>%
                           group_by(STATECD_NWOS) %>%
                           summarize(ACRES_PREV_NONFOREST = sum(ACRES), .groups = "drop"),
                         estimates %>%
                           filter(PREV_COND_STATUS_CD == "NONFOREST" & !is.na(PREV_PRESNFCD)) %>%
                           group_by(STATECD_NWOS) %>%
                           summarize(ACRES_PREV_NONFOREST_NOT_MISSING = sum(ACRES), .groups = "drop"),
                         by = join_by(STATECD_NWOS)) %>%
  mutate(PREV_COND_STATUS_CD = "NONFOREST", 
         PREV_NF_ADJ = ACRES_PREV_NONFOREST / ACRES_PREV_NONFOREST_NOT_MISSING) %>%
  select(STATECD_NWOS, PREV_COND_STATUS_CD, PREV_NF_ADJ)
estimates <- estimates %>%
  filter(!(PREV_COND_STATUS_CD == "NONFOREST" & is.na(PREV_PRESNFCD))) %>%
  left_join(prev.nf.adj, 
            by = join_by(STATECD_NWOS, PREV_COND_STATUS_CD)) %>%
  mutate(ACRES = if_else(is.na(PREV_NF_ADJ), ACRES, ACRES * PREV_NF_ADJ)) %>%
  select(-PREV_NF_ADJ)
```

```{r Reassess Missingness Re-check}
estimates %>% filter(COND_STATUS_CD == "NONFOREST") %>% count(LU)
estimates %>% filter(PREV_COND_STATUS_CD == "NONFOREST") %>% count(PREV_LU)

estimates %>% count(COND_STATUS_CD) 
estimates %>% count(PREV_COND_STATUS_CD)

estimates %>% filter(COND_STATUS_CD == "NONFOREST") %>% count(PRESNFCD)
estimates %>% filter(PREV_COND_STATUS_CD == "NONFOREST") %>% count(PREV_PRESNFCD)

estimates %>% filter(COND_STATUS_CD == "NONFOREST") %>% count(LU) 
estimates %>% filter(PREV_COND_STATUS_CD == "NONFOREST") %>% count(PREV_LU)
```

## Re-check Total Forest Acreage 
```{r Forest Acreage Re-check}
left_join(estimates %>%
            filter(COND_STATUS_CD == "FOREST") %>%
            group_by(STATECD_NWOS) %>%
            summarize(ACRES = sum(ACRES)),
          base.estimates %>%
            filter(YEAR == 2023 & COND_STATUS_CD == "FOREST") %>%
            group_by(STATECD_NWOS) %>%
            summarize(ACRES_BASE = sum(ACRES))) %>%
  mutate(DIFF = (ACRES - ACRES_BASE) / 1e6)
```

## Land Use Transition Matrix
```{r Land Use Transition Matrix}
lu.tran <- estimates %>%
  mutate(ACRES = ACRES / REMPER) %>%
  group_by(LU, PREV_LU) %>%
  summarize(ACRES = sum(ACRES) / 1e6, .groups = "drop") %>%
  mutate(ACRES = if_else(LU == PREV_LU, as.numeric(NA), ACRES)) %>%
  filter(!is.na(ACRES)) %>%
  mutate(across(c(LU, PREV_LU), ~ factor(.x, levels = c("FOREST", "AGRICULTURE",
                                                        "RANGELAND", "DEVELOPED", 
                                                        "OTHER")))) %>%
  pivot_wider(names_from = LU, values_from = ACRES) %>%
  arrange(PREV_LU) %>%
  select(PREV_LU, FOREST, AGRICULTURE, RANGELAND, DEVELOPED, OTHER)
lu.tran <- left_join(lu.tran,
                     lu.tran %>%
                       mutate(across(c(FOREST, AGRICULTURE, RANGELAND, DEVELOPED, OTHER), 
                                     ~ if_else(is.na(.x), 0, .x))) %>%
                       mutate(TOTAL = FOREST + AGRICULTURE + RANGELAND + DEVELOPED + OTHER) %>%
                       select(PREV_LU, TOTAL), by = join_by(PREV_LU)) %>%
  bind_rows(lu.tran %>%
              summarize(across(c(FOREST, AGRICULTURE, RANGELAND, DEVELOPED, OTHER), 
                               ~ sum(.x, na.rm = T))) %>%
              mutate(PREV_LU = "TOTAL")) %>%
  mutate(across(FOREST:TOTAL, ~ round(.x, 1)))
lu.tran
write_csv(lu.tran, "../DATA/LANDUSE_TRANSITION_MATRIX.csv")
```

## Gain/Loss Map
```{r Gain Loss Maps}
plot.loss <- plot %>%
  filter(PREV_COND_STATUS_CD == "FOREST" & COND_STATUS_CD != "FOREST") 
plot.loss <- usmap_transform(plot.loss, input_names = c("LON", "LAT")) %>%
  mutate(PRESNFCD = factor(PRESNFCD, 
                           levels = c("AGRICULTURE", "RANGELAND", "DEVELOPED", "OTHER"),
                           labels = c("Agriculture", "Rangeland", "Developed", "Other")))
plot_usmap("state") + 
  geom_sf(data = plot.loss, aes(color = PRESNFCD)) + 
  ggtitle("A. Forest loss") +
  scale_color_manual(values = c("#CC79A7", "#56B4E9", "#E69F00", "#999999"), 
                     name = "Lost to:") +
  theme(legend.position = "right")
ggsave("../FIGURES/FOREST_LOSS_MAP.pdf", width = 6, height = 3)
ggsave("../FIGURES/FOREST_LOSS_MAP.png", width = 6, height = 3)

plot.gain <- plot %>%
  filter(PREV_COND_STATUS_CD != "FOREST" & COND_STATUS_CD == "FOREST" & !is.na(PREV_PRESNFCD)) 
plot.gain <- usmap_transform(plot.gain, input_names = c("LON", "LAT")) %>%
  mutate(PREV_PRESNFCD = factor(PREV_PRESNFCD, 
                                levels = c("AGRICULTURE", "RANGELAND", "DEVELOPED", "OTHER"),
                                labels = c("Agriculture", "Rangeland", "Developed", "Other")))
plot_usmap("state") + 
  geom_sf(data = plot.gain, aes(color = PREV_PRESNFCD)) + 
  ggtitle("B. Forest gain") +
  scale_color_manual(values = c("#CC79A7", "#56B4E9", "#E69F00", "#999999"), 
                     name = "Gain from:") +
  theme(legend.position = "right")
ggsave("../FIGURES/FOREST_GAIN_MAP.pdf", width = 6, height = 3)
ggsave("../FIGURES/FOREST_GAIN_MAP.png", width = 6, height = 3)
```

## Forest Lost by Forest Type

*ADJUST FOR MISSINGNESS*

```{r PREV_FOREST_TYPE}
estimates %>% filter(PREV_COND_STATUS_CD == "FOREST") %>% count(PREV_FORTYPCD)

ft.adj <- left_join(estimates %>%
                      filter(PREV_COND_STATUS_CD == "FOREST") %>%
                      group_by(STATECD_NWOS) %>%
                      summarize(ACRES_FOREST = sum(ACRES), .groups = "drop"),
                    estimates %>%
                      filter(PREV_COND_STATUS_CD == "FOREST" & !is.na(PREV_FORTYPCD)) %>%
                      group_by(STATECD_NWOS) %>%
                      summarize(ACRES_FOREST_FT = sum(ACRES), .groups = "drop"),
                    by = join_by(STATECD_NWOS)) %>%
  mutate(PREV_COND_STATUS_CD = "FOREST", 
         FT_ADJ = ACRES_FOREST / ACRES_FOREST_FT) %>%
  select(STATECD_NWOS, PREV_COND_STATUS_CD, FT_ADJ)
estimates.ft <- estimates %>%
  filter(!(PREV_COND_STATUS_CD == "FOREST" & is.na(PREV_FORTYPCD))) %>%
  left_join(ft.adj, 
            by = join_by(STATECD_NWOS, PREV_COND_STATUS_CD)) %>%
  mutate(ACRES = if_else(is.na(FT_ADJ), ACRES, ACRES * FT_ADJ)) %>%
  select(-FT_ADJ)

estimates.ft %>% filter(PREV_COND_STATUS_CD == "FOREST") %>% count(PREV_FORTYPCD)

forest.loss.ft <- estimates.ft %>%
  mutate(ACRES = ACRES / REMPER) %>%
  left_join(ref.foresttype, by = join_by(PREV_FORTYPCD == FOREST_TYPE_CD)) %>%
  filter(PREV_LU %in% c("FOREST"), !LU %in% c("FOREST")) %>%
  group_by(FOREST_TYPE_GROUP) %>%
  summarize(ACRES = sum(ACRES) / 1e6, .groups = "drop") %>%
  mutate(PERC = (ACRES / sum(ACRES)) * 100,
         FOREST_TYPE_GROUP = gsub(" group", "", FOREST_TYPE_GROUP),
         FOREST_TYPE_GROUP = gsub(" / ", "/", FOREST_TYPE_GROUP)) %>%
  arrange((ACRES))
forest.loss.ft <- forest.loss.ft %>%
  mutate(FOREST_TYPE_GROUP = factor(FOREST_TYPE_GROUP, levels = forest.loss.ft$FOREST_TYPE_GROUP))

forest.loss.ft

ggplot(forest.loss.ft %>% slice_max(PERC, n = 10)) + 
  geom_bar(aes(x = FOREST_TYPE_GROUP,  y = PERC), stat = "identity", fill = "#a6611a") +
  coord_flip() +
  scale_y_continuous(expand = c(0, 0), lim = c(0, 25)) + 
  labs(x = element_blank(), y = "Percent of Forest Lost") +
  theme_linedraw() +
  theme(panel.grid.major.y = element_blank(),
        axis.ticks.y = element_blank())
ggsave("../FIGURES/FOREST_TYPE_LOSS.pdf", width = 5, height = 3)
ggsave("../FIGURES/FOREST_TYPE_LOSS.png", width = 5, height = 3)
```

## Forest Lost by Stand Origin
```{r PREV_STDORGCD}
estimates %>% filter(PREV_COND_STATUS_CD == "FOREST") %>% count(PREV_STDORGCD)

estimates %>%
  mutate(ACRES = ACRES / REMPER) %>%
  filter(PREV_LU %in% c("FOREST"), !LU %in% c("FOREST")) %>%
  group_by(PREV_STDORGCD) %>%
  summarize(ACRES = sum(ACRES) / 1e6, .groups = "drop") %>%
  arrange(desc(ACRES))  %>%
  mutate(PERCENT = ACRES / sum(ACRES))
```

## Forest Lost by Site Productivity
```{r PREV_SITECLCD}
estimates %>% filter(PREV_COND_STATUS_CD == "FOREST") %>% count(PREV_SITECLCD)

estimates %>%
  mutate(ACRES = ACRES / REMPER) %>%
  filter(PREV_LU %in% c("FOREST"), !LU %in% c("FOREST")) %>%
  mutate(PREV_SITECLCD_GRP = case_match(PREV_SITECLCD,
                                        c("1", "2", "3") ~ "HIGH",
                                        c("4", "5", "6") ~ "MODERATE",
                                        c("7") ~ "LOW")) %>%
  group_by(PREV_SITECLCD_GRP) %>%
  summarize(ACRES = sum(ACRES) / 1e6, .groups = "drop") %>%
  mutate(PERCENT = round((ACRES / sum(ACRES)) * 100, 1))
```

## Forest Lost by Ownership Group
```{r Ownership}
estimates %>% filter(PREV_COND_STATUS_CD == "FOREST") %>% count(PREV_OWNCD)

forest.loss.owncd <- estimates %>%
  mutate(ACRES = ACRES / REMPER) %>%
  filter(PREV_LU %in% c("FOREST"), !LU %in% c("FOREST")) %>%
  group_by(PREV_OWNCD) %>%
  summarize(ACRES = sum(ACRES) / 1e6, .groups = "drop") %>%
  arrange(desc(ACRES)) %>%
  mutate(PERCENT = (ACRES / sum(ACRES)) * 100) %>%
  mutate(PREV_OWNCD = factor(PREV_OWNCD, levels = rev(c("FAMILY", "CORPORATE", "FEDERAL", "STATE", "LOCAL", "TRIBAL")), 
                             labels = rev(c("Family", "Corporate", "Federal", "State", "Local", "Tribal"))))
forest.loss.owncd
ggplot(forest.loss.owncd %>% slice_max(PERCENT, n = 10)) + 
  geom_bar(aes(x = PREV_OWNCD,  y = PERCENT), stat = "identity", fill = "#a6611a") +
  coord_flip() +
  scale_y_continuous(expand = c(0, 0), lim = c(0, 60)) +
  labs(x = element_blank(), y = "Percent of Forest Lost") +
  theme_linedraw() +
  theme(panel.grid.major.y = element_blank(),
        axis.ticks.y = element_blank())
ggsave("../FIGURES/FOREST_OWN_LOSS.pdf", width = 5, height = 3)
ggsave("../FIGURES/FOREST_OWN_LOSS.png", width = 5, height = 3)
```

# Literature Cited
Burrill, E. A., A. M. DiTommaso, J. A. Turner, S. A. Pugh, G. Christensen, C. J. Perry, L. C. Lepine, D. M. Walker, and B. L. Conkling. 2024. The Forest Inventory and Analysis Database, FIADB user guides, volume database description (version 9.2), nationwide forest inventory (NFI). U.S. Department of Agriculture, Forest Service, Washington, DC. Available online at: https://research.fs.usda.gov/understory/forest-inventory-and-analysis-database-user-guide-nfi; last accessed September 4, 2024.

Pugh, S. A., J. A. Turner, E. A. Burrill, and W. David. In review. The Forest Inventory and Analysis Database: Population estimation user guide (Edition: 2024). U.S. Department of Agriculture, Forest Service.

# Appendices

## Metadata
Run date: `r Sys.Date()`

EVALIDs: `r paste(evalgrps %>% pull(EVAL_GRP), collapse = ", ")`

## Functions and Scripts
*NEED TO UPDATE*
```{r get_evalids.R}
get_evalgrps
```
```{r get_evalids.sql}
read.delim("SQL/get_evalgrps.sql", header=F)
```

```{r get_counties.R}
get_counties
```
```{r get_forestarea.R}
get_forestarea
```

```{r get_forestarea.sql}
read.delim("SQL/get_forestarea.sql", header=F)

```